\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-02-22 papers/methods/paper.tex
Paper updates

The first layer of the hierarchy represents the latent population of $m_B$, $x_1$ and $c$. This can be thought of as the `true' values for the supernova, and with correct summary statistics and uncertainty would be normally distributed around the observed values such that 
\begin{equation}
P(\hat{m_B},\hat{x_1},\hat{c}|m_B, x_1, c,\cov) = \mathcal{N}(\lbrace \hat{m_B},\hat{x_1},\hat{c} \rbrace|\lbrace m_B, x_1, c \rbrace, \cov). \label{eq:pop}
\end{equation}
However, the presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics report from light curve fitting are not Gaussian. We model the extra dispersion included in the observations from colour smearing by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra dispersion shows heavy redshift dependence. As this may represent underlying physics more than the effects of Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $(\kappa_0 + \kappa_1 z)^2$ to our observed colour uncertainty, where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. Each survey has its own set of $\kappa$ parameters. Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the bias in the mean colour measured as a function of redshift, and subtract this bias from our observed data.

@@

Each realisation of cosmology fitted contains 137 LowZ supernovae, and 204 DES-like supernovae, such that the uncertainties found when combining chains is representative of the uncertainty in the final DES spectroscopic analysis. {\red Focus on w}.

 Combined posterior surfaces for 100 data realisations are shown in Figure \ref{fig:bulk_summary}. Note that we have combined the posteriors for 100 realisations, and so we should expect the size of the uncertainty to be representative of one realisation, but the statistical spread of the final surface should be $\sqrt{100} = 10$ times less than a single realisation. Plot summaries for the fits are shown in Figure \ref{fig:bulk_summary}, and the parameter bounds are listed in Table \ref{tab:bulk_summary}. The results indicate changes in the underlying population can effect significant changes in the recovered standardisation parameters $\alpha$ and $\beta$, however do not appear to have a significant effect on recovered cosmology. Whilst the combined sample (on which the selection function was fit) shows no significant bias (as expected), the {\celeven} and {\gten} only models show opposite biases in $\Omega_m$. Cosmological bias is detected with $\approx 3.0\sigma$ significance for the {\celeven} model and $\approx 1.7\sigma$ significance for the {\gten} model, indicating that the population and selection effect treatment cannot capture all necessary information to encapsulate the magnitude and chromatic smearing of the supernovae population. However, for the sample size of the DES and LowZ supernova samples (of order 350 supernova), these effects are sub-dominant to the statistical uncertainty, representing at most a deviation of $0.25\sigma$ and, accounting for statistical fluctuations up to $2\sigma$, a minimum bias of $0.08\sigma$ for the {\celeven} model.

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk2.pdf}
	\end{center}
	\caption{Graphical parameter summaries for 100 realisations of supernova data.}
	\label{fig:bulk_summary}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_small2.pdf}
	\end{center}
	\caption{Posterior surfaces for 100 realisations of supernova data. Neither population nor scatter model has a significant effect on posterior surface location or shape. Population skewness has the primary effect of changing the degeneracy direction for the $\alpha$ -- $\beta$ contour, whilst the scatter models and skewness have noticeable effects on the value and uncertainty of the recovered parameter $\beta$.}
	\label{fig:bulk_posterior}
\end{figure}

@@

As stated in Section \ref{sec:selection}, the underlying population in apparent magnitude, when we discard skewness, can be represented as $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where
\begin{align}
m_B^*(z) &= \langle M_B \rangle + \mu(z) - \alpha \langle x_1(z) \rangle + \beta \langle c(z) \rangle\label{eq:disc1} \\
\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 + (\beta \sigma_c)^2 \notag\\ 
&\quad\quad + 2(\beta \sigma_{M_B,c} - \alpha \sigma{m_B,x_1} - \alpha \beta \sigma_{x_1,c}). \label{eq:disc2}
\end{align}
Then, modelling $P(S|m_B)$ as either a normal or a skew normal, we can analytically perform the integral and reach equations \eqref{eq:seldes} and \eqref{eq:sellowz}.

>>>>

The first layer of the hierarchy represents the latent population of $m_B$, $x_1$ and $c$. This can be thought of as the `true' values for the supernova, and with correct summary statistics and uncertainty would be normally distributed around the observed values such that 
\begin{equation}
P(\hat{m_B},\hat{x_1},\hat{c}|m_B, x_1, c,\cov) = \mathcal{N}(\lbrace \hat{m_B},\hat{x_1},\hat{c} \rbrace|\lbrace m_B, x_1, c \rbrace, \cov). \label{eq:pop}
\end{equation}
However, the presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics report from light curve fitting are not Gaussian. We model the extra dispersion included in the observations from colour smearing by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra dispersion shows heavy redshift dependence. As this may represent underlying physics more than the effects of Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. Each survey has its own set of $\kappa$ parameters. Furthermore, we use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the bias in the mean colour measured as a function of redshift, and subtract this bias from our observed data. We allow the amount of bias to be parametrised to account for different intrinsic scatter models, where parameter $\kappa_2$ controlled the fraction of bias subtraction - 0 for no bias, up to 1 for the maximum bias found in simulations (which corresponded to the intrinsic scatter model from \citet{Chotard2011}). Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $\kappa_2$ represents a universal shift in supernovae biases. We place a uniform prior on $\kappa_2$.

@@

Each realisation of cosmology fitted contains 137 LowZ supernovae, and 204 DES-like supernovae, such that the uncertainties found when combining chains is representative of the uncertainty in the final DES spectroscopic analysis. As our primary focus is upon Dark Energy, we now focus specifically on the Flat $w$CDM model with matter prior.

 Combined posterior surfaces for 100 data realisations are shown in Figure \ref{fig:bulk_posterior}. Note that we have combined the posteriors for 100 realisations, and so we should expect the size of the uncertainty to be representative of one realisation, but the statistical spread of the final surface should be $\sqrt{100} = 10$ times less than a single realisation. The parameter bounds are listed in Table \ref{tab:bulk_summary}.

Cosmological bias is detected with $\approx 2.7\sigma$ significance for the {\celeven} model and $\approx 1.5\sigma$ significance for the {\gten} model when fitting both systematics and statistics. For the statistics only fits, the respective bias detection levels are $0.9\sigma$ and $5.5\sigma$. We investigate the source of the {\celeven} bias and find its source to be bias in the observed summary statistics, in addition to incorrect reported uncertainty on those summary statistics. By replacing fully simulated observed $m_B$, $x_1$ and $c$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. Unfortunately, adding extra fit parameters to allow for shifting observables washes out cosmology, and applying a specific bias correction requires running a fiducial simulation (assuming cosmology, population and scatter model) and binning data, and is difficult to do whilst accounting for correlations with population and scatter model. This is compounded by the fact that bias corrections do not in general improve fits (increase the log posterior), and so are difficult to fit inherently. Colour bias corrections for the {\celeven} scatter model are included in our cosmology model, the strength of which are controlled by a fit parameter, $\kappa_2$, however neither the {\gten} nor {\celeven} simulations constrained the fitting of this parameter, and thus the majority of bias in the {\celeven} fits remain. Works such as \citet{Kessler2017} show that bias corrections can be applied to supernovae datasets that can robustly handle multiple intrinsic scatter models, and future work will center on uniting these methodologies - incorporating better bias corrections without having to precompute standardisation parameters and populations.

For the sample size of the DES and LowZ supernova samples (of order 350 supernova), the bias from intrinsic scatter models are sub-dominant to the statistical uncertainty, representing at most a deviation of $0.27\sigma$ for the full statistics and systematics model fit for the {\celeven} model, and as such we will leave more complicated treatment of them for future work.

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_small2.pdf}
	\end{center}
	\caption{Posterior surfaces for 100 realisations of supernova data. Surfaces for $\Omega_m$, $w$, $\alpha$, $\beta$ and $\langle M_B \rangle$ are shown, with the other fit parameters being marginalised over. The intrinsic scatter model has significant impact on the recovered $\beta$, which in itself effects cosmology, resulting in small biases in $w$.}
	\label{fig:bulk_posterior}
\end{figure}

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. Whilst the {\gten} models do not show significant evidence of bias, the {\celeven} model does. With the current statistical size of the supernovae dataset, the bias is sub-dominant to the size of the uncertainty. }
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ & Bias Detection Level \\
		\hline
		{\gten} Stat + Syst  &  $-1.02\pm 0.13$ & $1.5\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.97\pm 0.11$ & $2.7\sigma$  \\
		{\gten}  Stat         &  $-1.01\pm 0.11$ & $0.9\sigma$  \\
		{\celeven} Stat         &  $-0.95\pm 0.09$ & $5.5\sigma$  \\
		\hline
	\end{tabular}
\end{table}

@@

As stated in Section \ref{sec:selection}, the underlying population in apparent magnitude, when we discard skewness, can be represented as $\mathcal{N}(m_B|m_B^*(z), \sigma^*_{m_B})$, where
\begin{align}
m_B^*(z) &= \langle M_B \rangle + \mu(z) - \alpha \langle x_1(z) \rangle + \beta \left(\langle c(z) \rangle + \sqrt{\frac{2}{\pi}}\sigma_c \delta_c\right)\label{eq:disc1} \\
\sigma^*_{m_B} &= \sigma_{M_B}^2 + (\alpha \sigma_{x_1})^2 +  \left(\beta \sigma_c \sqrt{1 - \frac{2\delta_c^2}{\pi}}\right)^2. \label{eq:disc2}
\end{align}
Then, modelling $P(S|m_B)$ as either a normal or a skew normal, we can analytically perform the integral and reach equations \eqref{eq:seldes} and \eqref{eq:sellowz}.
